On the power of conditional independence testing \\ under model-X

### 2022-02-28 arxiv-v4/response_to_reviews.tex

\documentclass[12pt]{article}
\usepackage{amssymb,amsfonts,amsmath, bm}
\oddsidemargin=.08in
\evensidemargin=.08in
\textwidth=6.2in
\topmargin=-.5in
\textheight=9in
\usepackage{url}
%\usepackage[authoryear]{natbib}
\usepackage{hyperref}
\date{\today}

\newcommand{\Ids}{\mathds{I}}
\newcommand{\Pds}{\mathds{P}}
\newcommand{\Eds}{\mathds{E}}
\newcommand{\Vds}{\mathds{V}}
\newcommand{\Ical}{\mathcal{I}}
\newcommand{\Rcal}{\mathcal{R}}
\newcommand{\Ccal}{\mathcal{C}}
\newcommand{\Acal}{\mathcal{A}}

%\def\T{{ \mathrm{\scriptscriptstyle T} }}
%\def\v{{\varepsilon}}
%%\def\ibinom#1#2{{C(#1,#2)}}
%\def\ibinom#1#2{{\binom{#1}{#2}}}

\begin{document}

\title{Response to reviews of\\ ``On the power of conditional independence testing \\ under model-X"\\ \vskip.2cm \Large EJS2107-037}

\author{Eugene Katsevich and Aaditya Ramdas}

\maketitle

We would like to thank the referee and the associate editor for useful comments and for the opportunity to revise our paper. Based on your suggestions, we made several improvements to the revised manuscript. We summarize these next, and then provide point-by-point responses to the comments.

\section{Overview of Revisions}

\paragraph{}

Next, we give our point-by-point responses.

\section{Responses to Referee}

\begin{enumerate}
	\item \textsl{I do find the different sections and aims of the paper a bit disconnected— specifically I’m not sure if (1) is well-connected with (2) and (3). (1) is to do with the power of the CRT test. On the other hand, (2) is about a different test that works with different assumptions, though there is an asymptotically equivalent CRT test, and (3) studies the power of this test. The motivation behind (2) seems to be concerns about the validity of the CRT assumptions (which require full knowledge of the conditional distribution $X|Z$), but I don’t see how this idea of ‘robust- ness’ really relates closely to (1). One option could be to remove or de-emphasize (1)—in my opinion it is nice to have such a result written down somewhere, but it is unsurprising.}\\

	We thank the referee for pointing out the somewhat disconnected nature of the submitted manuscript, especially the transition from Section 2 to Section 3. We agree, and we have substantially edited the exposition in Section 3 for a smoother flow. The high-level logic in the revised manuscript is that Section 2 studies the power and optimality of the CRT in finite samples against point alternatives while Sections 3 and 4 study the power of the CRT in an asymptotic setting against semiparametric alternatives. Section 3 prepares the groundwork by establishing the asymptotic equivalence between the distilled CRT and the simpler-to-analyze MX(2) $F$-test (loosening the assumption from MX to MX(2) along the way), while Section 4 establishes the power of the MX(2) $F$-test and therefore the distilled CRT. Ultimately, the central goal of the manuscript is to study the power and optimality of the CRT, and the MX(2) $F$-test is mostly a means to that end---though we believe it is of independent interest as well. This logic is conveyed in the opening paragraph of Section 3:

	\begin{quote}
		``In Section 2, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\bm Y| \bm X, \bm Z}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this ML procedure. Specifically, we consider  the power of a specific instance of the CRT against a set of semiparametric alternatives (Section 3.1). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section 3.2), which is of independent interest due to its closed form and weaker assumptions (Section 3.3). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section 3.4) and put this section's results into perspective (Section 3.5) before moving on to stating the desired power results in the next section (Section 4).''
	\end{quote}

	\item \textsl{Regarding the robustness of the CRT test, I feel one could argue that assuming exact knowledge of the first two conditional moments of $X | Z$ is not too much weaker than asking for knowledge of the entire distribution. Are there specific examples where completely correct models for the conditional means and variances are known, but not the conditional distribution? In my opinion, a more relevant analysis would quantify the impact of the error in estimating these conditional moments. Also, it might be interesting to consider a version that does not require estimating the second conditional moment, but potentially places a slightly stronger condition on the error in estimating $\mathbb E[Y | Z]$.}\\

	We agree with the referee that assuming exact knowledge of the first two moments of $X|Z$ is still a strong assumption, which we would not necessarily expect to be satisfied in practice. We nevertheless find it comforting that the MX methodologies considered here are not sensitive to misspecification of the higher moments of the assumed $X|Z$. We agree that a stronger kind of robustness analysis would quantify the impact of the error in estimating these conditional moments and/or shift more of the assumptions onto $\mathbb E[Y|Z]$, but these directions are beyond the scope of the current work. We convey these points in Section 3.5 of the revised manuscript:
	\begin{quote}
	``...Theorem 3 shows that asymptotic Type-I error control of MX-style methodologies can be achieved under the weaker MX(2) assumption, requiring only two moments of the conditional distribution $\mathcal L_n(\bm X | \bm Z)$ rather than the entire conditional distribution. If strict Type-I error control in finite samples is desired, however, then we must continue to rely on the full MX assumption. Nevertheless, our simulation suggests the MX(2) $F$-test has fairly good Type-I error control in finite samples as well. Finally, note that the MX(2) assumption still requires \textit{exact} knowledge of the first and second conditional moments; we leave as an important future direction to examine the robustness of these tests to errors in these quantities (bounds on the worst-case Type-I error of the CRT have been obtained in the context of a misspecified $\mathcal L_n(\bm X | \bm Z)$ (Berrett et al 2020)).''
	\end{quote}

	\item \textsl{What is the significance of the asymptotic equivalence of the CRT test constructed using the test statistic on which the MX(2) $F$-test is based, and the MX(2) $F$-test itself? I cannot see why the CRT test would be preferred here? Overall, the connection between CRT tests and the MX(2) $F$-test seems quite weak to me.} \\

	The significance is that we can study the power of the CRT by instead studying the power of the easier-to-analyze MX(2) $F$-test. We have clarified this motivation to construct the MX(2) $F$-test in the first paragraph of Section 3 (quoted above). Furthermore, we expand on the methodological advantages and disadvantages of the MX(2) $F$-test compared to the CRT in Section 3.5 of the revised manuscript:
	\begin{quote}
		``The preceding results suggest that the MX(2) $F$-test is a useful alternative to the dCRT: the power of these methods is asymptotically the same (Theorem 2), while the MX(2) $F$-test is computationally faster because it does not require resampling (Table~\ref{tab:summary-1}). On the other hand, note that we have proven Type-I error control for the MX(2) $F$-test only when $\widehat g_n$ is fit out of sample and only asymptotically, while the dCRT gives finite-sample Type-I error with in-sample fit $\widehat g_n$. However, numerical simulations suggest good finite-sample Type-I error control for the MX(2) $F$-test even when $\widehat g_n$ is fit in sample.''
	\end{quote}

	\item \textsl{The methodological contribution here is quite modest as the proposed MX(2) F-test is essentially a multivariate version of the Generalized covariance measure proposed by Shah \& Peters.}
\\

We view our contribution as primarily theoretical, establishing basic power and optimality results for increasingly popular methodologies like the CRT. Furthermore, we agree that the MX(2) $F$-test has a close relationship to the GCM test. Nevertheless, the MX(2) $F$-test complements the GCM test in the sense that it controls Type-I error in regimes when the latter is not guaranteed to, e.g. in cases when $\mathbb E[\bm Y | \bm Z]$ cannot be consistently estimated. We compare the two methodologies in Section 3.5 of the revised manuscript and in Tables~\ref{tab:summary-1} and~\ref{tab:summary-2}, reproduced here:

\begin{quote}
``To compare our results with those in non-parametric doubly robust inference, we consider the closest representative of the latter: the generalized covariance measure (GCM) test of Shah and Peters. The GCM test statistic is defined as
\begin{equation}
	\widehat \rho^{\text{GCM}}_n \equiv \frac{1}{n}\sum_{i = 1}^n  (Y_i - \widehat g_n(Z_i))(X_i - \widehat \mu_n(Z_i)),
\end{equation}
where $\widehat \mu_n(\bm Z)$ and $\widehat g_n(\bm Z)$ are estimates of $\mu_n(\bm Z) = \mathbb E_{\mathcal L_n}[\bm X|\bm Z]$ and $g_n(\bm Z) \equiv \mathbb E_{\mathcal L_n}[\bm Y|\bm Z]$, respectively. This statistic is shown to converge under conditional independence to a mean-zero normal limit as long as the estimates of $\mathbb E_{\mathcal L_n}[\bm X|\bm Z]$ and $\mathbb E_{\mathcal L_n}[\bm Y|\bm Z]$ are both consistent, while the product in these estimation errors tends to zero at a rate of $o(n^{-1/2})$. By contrast, the MX(2) $F$-test places more weight on the model for $\bm X|\bm Z$ (assuming both first and second moments of this conditional distribution are known) while placing less weight on the model for $\bm Y|\bm Z$ (not assuming even consistency for $\mathbb E_{\mathcal L_n}[\bm Y|\bm Z]$). Therefore, while the MX(2) $F$-test closely resembles the GCM test, the assumptions required for validity of these two methods do not subsume each other (Table~\ref{tab:summary-2}).''
\end{quote}

	\begin{table}[h!]
	\centering
	\begin{tabular}{l|ll}
		Method & Guarantee & Resampling \\
		\hline
		CRT & Finite-sample & Yes \\	
		MX(2) $F$-test & Asymptotic & No  \\
		GCM test & Asymptotic & No
	\end{tabular}
	\caption{Type-I error guarantee and necessity of resampling for each method compared.}
	\label{tab:summary-1}
\end{table}

\begin{table}[h!]
	\small
	\begin{tabular}{l|lllll}
		Method & $\mathcal E(\mathbb E[\bm X|\bm Z]$) & $\mathcal E(\text{Var}[\bm X|\bm Z])$ & $\mathcal E(\mathcal L(\bm X|\bm Z))$ & $\mathcal E(\mathbb E[\bm Y|\bm Z]$) & $\mathcal E(\mathbb E[\bm X|\bm Z]) \times \mathcal E(\mathbb E[\bm Y|\bm Z])$ \\
		\hline
		CRT & 0 & 0 & 0 & --& --  \\
		MX(2) $F$-test  & 0 & 0 & -- & -- & -- \\
		GCM test & $o(1)$ & -- & -- & $o(1)$ & $o(n^{-1/2})$
	\end{tabular}
	\caption{Assumptions necessary for each method compared (excluding moment assumptions). Here, $\mathcal E(\cdot)$ refers to the root-mean-squared estimation error of a given quantity.}
	\label{tab:summary-2}
\end{table}

\end{enumerate}

\end{document}

### 2022-02-28 arxiv-v4/crtpower.tex
Added draft of arXiv-v4

 \usepackage[vmargin=1.18in,hmargin=1.1in]{geometry}
 \usepackage{amssymb,amsfonts,graphicx,amsthm,nicefrac,mathtools,bm, bbm}
 \usepackage[hidelinks]{hyperref}
 \usepackage[normalem]{ulem} %to strike the words
 \usepackage{mathrsfs}

@@

\addbibresource{~/.references/crtpower}
%\addbibresource{crtpower}

@@

\section{Weakening the MX assumption to MX(2)} \label{sec:weakening}

Instead of assuming knowledge of the entire conditional distribution $\prx|\prz$, assume only 
\begin{equation}
\text{the conditional mean } \mathbb E[\prx|\prz] \text{ and variance } \text{Var}[\prx|\prz] \text{ are known}.
\label{mx-assumption}
\end{equation}
We call this the \textit{MX(2) assumption}. In this section, we show that asymptotic Type-I error can be uniformly controlled under this drastically weaker assumption (together with moment assumptions). We work in an asymptotic regime described by Setting~\ref{setting:asymptotic} below.

\begin{setting}[\bf Arbitrary dimension asymptotics] \label{setting:asymptotic}

Following a triangular array formalization, for each $n = 1, 2, \dots$, we have a joint law $\mathcal L_n$ over $(\prx, \pry, \prz) \in \mathbb R^{d + r + p}$, where $d = \text{dim}(\prx)$ remains fixed, $r = \text{dim}(\pry) = 1$, and $p = \text{dim}(\prz)$ can vary arbitrarily with $n$. Under the MX(2) assumption, the law $\mathcal L_n$ has known mean and variance functions
\begin{equation}
	\mu_n(\prz) \equiv \mathbb E_{\mathcal L_n}[\prx|\prz]  \text{ and } \Sigma_n(\prz) \equiv \text{Var}_{\mathcal L_n}[\prx|\prz].
	\label{conditional-mean-variance}
\end{equation}
For each $n$, we receive $n$ i.i.d. samples $(\srx, \sry, \srz) = \{(\srx_i, \sry_i, \srz_i)\}_{i = 1}^n$ from $\mathcal L_n$. Note that we leave implicit the dependence on $n$ of $(\prx,\pry,\prz)$ and $(X,Y,Z)$ to lighten the notation.

\end{setting}

By analogy with definition~\eqref{eq:null-under-modelX}, the MX(2) null hypothesis is defined as 
\begin{equation}
	\mathscr L_0^{\text{MX(2)}} = \mathscr L_0^{\text{MX(2)}}(\mu_n(\cdot), \Sigma_n(\cdot)) \equiv \mathscr L_{0} \cap \mathscr L^{\text{MX(2)}}(\mu_n(\cdot), \Sigma_n(\cdot)),
	\label{mx2-null}
\end{equation}
where
\begin{equation*}
	\begin{split}
		\mathscr L^{\textnormal{MX(2)}}(\mu_n(\cdot), \Sigma_n(\cdot)) \equiv \{\mathcal L_n: \mathbb E_{\mathcal L_n}[\prx|\prz] = \mu_n(\prz),\ \text{Var}_{\mathcal L_n}[\prx|\prz] = \Sigma_n(\prz)\}.
	\end{split}
\end{equation*}
In Section~\ref{sec:mx2-f-test}, we propose a test of this MX(2) null hypothesis called the \textit{MX(2) $F$-test}. We establish in Section~\ref{sec:MX(2)-regression} that this test controls Type-I error uniformly over subsets of $\mathscr L_0^{\text{MX(2)}}$ satisfying moment conditions and that this test is asymptotically equivalent to the CRT based on the same test statistic. We compare these results to existing ones in Section~\ref{sec:comparison-to-existing-results-3}, and then check their finite-sample accuracy via numerical simulations in Section~\ref{sec:simulations}. All proofs are deferred to Appendix~\ref{sec:proofs-sec45}.

\subsection{The MX(2) $F$-test} \label{sec:mx2-f-test}

Suppose we have trained an estimate $\widehat g_n$ of $\mathbb E_{\mathcal L_n}[\pry|\prz]$ on an independent dataset (whose size can vary arbitrarily with $n$ and is not included in the sample size $n$ used for testing). In the next section, $g_n$ will denote be the nonparametric portion of a semiparametric model~\eqref{eq:linearity}. These training sets across $n$ and resulting estimates $\widehat g_n$ remain fixed throughout. Importantly, the sample used for training $\widehat g_n$ \emph{need not come from the same distribution} as $\mathcal L_n$ and can therefore be much larger. For example, in the context of the digital twin test \cite{Bates2020} it was pointed out that large observational datasets can be used to train $\widehat f_{\pry|\prx,\prz}$, while applying a variant of the CRT to a smaller (randomized) experimental dataset to obtain a causal guarantee. Hence, our setting could be viewed from the lens of sample splitting, or in the spirit of drawing inferences when having a (typically large) observational dataset and a (likely small) experimental dataset. Both viewpoints will be useful in later sections.

With the estimate $\widehat g_n$ in hand, it is natural to base inference on the sample covariance between $\prx$ and $\pry$ after adjusting for $\prz$:
\begin{equation}
\widehat \rho_n \equiv \frac1n\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)).
\label{rho-hat}
\end{equation}
In general, $\widehat \rho_n \in \mathbb R^d$, but for $d = 1$, this coincides with the \textit{generalized covariance measure}, proposed by Shah and Peters \cite{Shah2018} for conditional independence testing. Related quantities also have been studied in the semiparametric \cite{Robinson1988, Robins1992} and doubly robust \cite{Robins2001, VanderLaan2003, Chernozhukov2018} estimation contexts; see Section~\ref{sec:comparison-to-existing-results-3} for more discussion. The quantity $\widehat \rho_n$ can be viewed as an estimator of $\rho_n \equiv \mathbb E_{\mathcal L_n}[\textnormal{Cov}_{\mathcal L_n}[\prx,\pry|\prz]]$, and in fact is a consistent estimator under the MX(2) assumption (see Lemma~\ref{lem:consistency}). Constructing an asymptotically valid CI test based on $\widehat \rho_n$ requires us to be able to consistently estimate the limiting variance of this quantity under the null. If we have access to the first two moments of $\prx|\prz$, we can compute for any $\mathcal L_n \in \mathscr L_0^{\text{MX(2)}}$ that
\begin{equation}
\begin{split}
\text{Var}_{\mathcal L_n}[\sqrt{n}\widehat \rho_n] &=\text{Var}_{\mathcal L_n}[(\pry - \widehat g_n(\prz))(\prx - \mu_n(\prz))] \\
&=\text{Var}_{\mathcal L_n}\left[\mathbb E_{\mathcal L_n}[(\pry - \widehat g_n(\prz))(\prx - \mu_n(\prz))|\pry,\prz]\right] + \\
&\quad \quad \mathbb E_{\mathcal L_n}\left[\text{Var}_{\mathcal L_n}[(\pry - \widehat g_n(\prz))(\prx - \mu_n(\prz))|\pry,\prz]\right] \\
&\quad = \mathbb E_{\mathcal L_n}\left[(\pry - \widehat g_n(\prz))^2 \Sigma_n(\prz)\right] \\
&\quad\equiv S^2_{n} \in \mathbb R^{d \times d}.
\label{variance-calculation}
\end{split}
\end{equation}
A natural estimate of this limiting variance is
\begin{equation}
\widehat S_n^2 \equiv \frac{1}{n}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))^2\Sigma_n(\srz_i).
\end{equation}
Note that we can compute both $\widehat S_n^2$ and $\widehat \rho_n$ using only the MX(2) assumption.

Based on the variance calculation~\eqref{variance-calculation}, we would expect the standardized quantity 
\begin{equation}
	U_n(\srx, \sry, \srz) \equiv \widehat S_n^{-1} \sqrt{n} \widehat \rho_n = \frac{\widehat S_n^{-1}}{\sqrt{n}}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)) \in \mathbb R^d.
	\label{u-hat}
\end{equation}
to converge to $N(0,I_d)$ under the MX(2) null~\eqref{mx2-null}. This motivates the \textit{MX(2) $F$-test} 
\begin{equation*}
\phi^{\text{MX(2)}}_n(\srx, \sry, \srz) \equiv \mathbbm 1(T_n(\srx, \sry, \srz) > c_{d,1-\alpha}),
\end{equation*}
where
\begin{equation}
T_n(\srx, \sry, \srz) \equiv \|U_n(\srx, \sry, \srz)\|^2.
\label{t-n}
\end{equation}
See also Algorithm~\ref{alg:MX(2)-F-test}, and recall that $c_{d,1-\alpha}$ is defined as the $1-\alpha$ quantile of $\chi^2_d$. 
\begin{center}
	\begin{minipage}{\linewidth}
		\begin{algorithm}[H]
			%			\SetAlgorithmName{Procedure}{}\; %last arg is the title of listing table		
			\KwData{
				$\{(\srx_i, \sry_i, \srz_i)\}_{i=1}^n$, $\mu_n(\cdot)$ and $\Sigma_n(\cdot)$ in \eqref{conditional-mean-variance}, learning method $g$
			}
			Obtain $\widehat g$ by fitting $g$ on  a separate dataset\;
			Recall $\mu_n(\srz_i) \equiv \mathbb E_{\mathcal L_n}[\srx_i|\srz_i], \text{ set } \widehat S_n^2 \equiv \frac{1}{n}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))^2\Sigma_n(\srz_i)$\;
			Set $\smash{U_n \equiv \frac{\widehat S_n^{-1}}{\sqrt{n}}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i))}$ and $T_n = \|U_n\|^2$\;
			% Compute $T(\srx,\sry,\srz) \equiv \sum_{i = 1}^n \ell(\sry_i, (\srx_i-\mu_n(\srz_i)) \widehat \beta + \widehat g_n(\srz_i))$\;
			% \For{$b \in \{1,\dots,B\}$}{
			% 	Sample $\srxk_i^b$ from $\mathcal L(\srx_i|\srz_i)$ for all $i$\;
			% 	Compute $T(\srxk^b,\sry,\srz)$ using steps 3, 4 with $\srxk_i^b$ replacing $\srx_i$\;
			% }
			% Compute $\widehat p_j$ based on formula~\eqref{CRT}\;
			\KwResult{MX(2) $F$-test asymptotic $p$-value $\widehat p \equiv \mathbb P[\chi^2_d > T_n]$.}
			\textbf{Cost:} One $p$-dimensional model fit.
			\caption{\bf The MX(2) $\bm F$-test}
			\label{alg:MX(2)-F-test}
		\end{algorithm}
	\end{minipage}
\end{center}
Note that a one-sided version of this test (the \textit{MX(2) $t$-test}) can be defined for $d = 1$ by rejecting for large values of $U_n(\srx, \sry, \srz)$.

Next, we formally state asymptotic type-I error control for the MX(2) $F$-test and claim that it is asymptotically equivalent to the CRT based on the same test statistic.

\subsection{Asymptotic type-I error control and equivalence to CRT} \label{sec:MX(2)-regression}

For uniform type-I error control, we must restrict the set of null distributions~\eqref{mx2-null} to those satisfying the following moment conditions for fixed $c_1, c_2 > 0$:
\begin{equation}
	\mathscr L_n(c_1, c_2) \equiv \{\mathcal L_n: \|S_{n}^{-1}\| \leq c_1, \mathbb E_{\mathcal L_n}\left[(\pry - \widehat g_n(\prz))^{4} \mathbb E_{\mathcal L_n}[\|\prx - \mu_n(\prz)\|^{4}|\prz]\right] \leq c_2 \}.
\end{equation} 
For pointwise error control (less desirable), these moment conditions can be relaxed, but they are unavoidable for uniform type-1 error control.

>>>>

 \usepackage[vmargin=1.18in,hmargin=1.1in]{geometry}
 \usepackage{amssymb,amsfonts,graphicx,amsthm,nicefrac,mathtools,bm, bbm}
 \usepackage[hidelinks]{hyperref}
 \usepackage[normalem]{ulem} %to strike the words
 \usepackage{mathrsfs}
 \usepackage{pdflscape}

@@

% \addbibresource{~/.references/crtpower}
\addbibresource{crtpower}

@@

\section{An asymptotic equivalent to the distilled CRT} \label{sec:weakening}

In Section~\ref{sec:power}, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\pry|\prx,\prz}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this ML procedure. Specifically, we consider  the power of a specific instance of the CRT against a set of semiparametric alternatives (Section~\ref{sec:asymptotic-setup}). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section~\ref{sec:second-order-approx}), which is of independent interest due to its closed form and weaker assumptions (Section~\ref{sec:mx2-f-test}). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section~\ref{sec:simulations}) and put this section's results into perspective (Section~\ref{sec:comparison-to-existing-results-3}) before moving on to stating the desired power results in the next section (Section~\ref{sec:asymptotic-power}).

\subsection{Semiparametric alternatives and the distilled CRT} \label{sec:asymptotic-setup}

First, we define an asymptotic framework within which we will work in Sections~\ref{sec:weakening} and~\ref{sec:asymptotic-power}. Following a triangular array formalization, for each $n = 1, 2, \dots$, we have a joint law $\mathcal L_n$ over $(\prx, \pry, \prz) \in \mathbb R^{d + r + p}$, where $d = \text{dim}(\prx)$ remains fixed, $r = \text{dim}(\pry) = 1$, and $p = \text{dim}(\prz)$ can vary arbitrarily with $n$. For each $n$, we receive $n$ i.i.d. samples $(\srx, \sry, \srz) = \{(\srx_i, \sry_i, \srz_i)\}_{i = 1}^n$ from $\mathcal L_n$. Note that we leave implicit the dependence on $n$ of $(\prx,\pry,\prz)$ and $(X,Y,Z)$ to lighten the notation. In this framework, it will be useful to define the mean and variance functions
\begin{equation}
	\mu_n(\prz) \equiv \mathbb E_{\mathcal L_n}[\prx|\prz]  \text{ and } \Sigma_n(\prz) \equiv \text{Var}_{\mathcal L_n}[\prx|\prz].
	\label{eq:conditional-mean-variance}
\end{equation}

Now, consider a set of semiparametric alternatives $\mathcal L_n(\pry|\prx,\prz)$ such that
\begin{equation}
	\pry = (\prx-\mu_n(\prz))^T \beta_n + g_n(\prz) + \peps; \quad \peps \sim N(0, \sigma^2 ),\ \sigma^2 > 0
	\label{eq:linearity}
\end{equation}
for $\peps \independent (\prx, \pry, \prz)$. Here, $\beta_n \in \mathbb R^d$ is a coefficient vector, $g_n: \mathbb R^p \rightarrow \mathbb R$ a general function, and $\sigma^2 > 0$ the residual variance. Of special interest are local alternatives where $\beta_n = h/\sqrt{n}$ for some $h \in \mathbb R^d$. By Theorem~\ref{thm:power}, the most powerful test against the alternative~\eqref{eq:linearity} is the CRT based on the likelihood statistic 
\begin{equation}
T^{\text{opt}}(\srx, \sry, \srz) = \prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T \beta_n - g_n(\srz_i)\right)^2\right).
\end{equation}
Assuming local alternatives $\beta_n = h/\sqrt n$ and taking a logarithm, we obtain
\begin{equation}
	\begin{split}
		\log T^{\text{opt}}(\srx, \sry, \srz) &= -\frac n 2 \log(2\pi) - \frac{1}{2\sigma^2}\sum_{i = 1}^n\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T h/\sqrt{n} - g_n(\srz_i)\right)^2 \\
		&\approx   \frac{h^T}{\sigma^2}\frac{1}{\sqrt{n}}\sum_{i = 1}^n(\sry_i - g_n(\srz_i))(\srx_i-\mu_n(\srz_i)) + C,
	\end{split}
\label{eq:optimal-semiparametric}
\end{equation}
where $C$ is a constant that does not depend on $\srx$ and therefore does not change upon resampling.

Of course, inference based on $T^{\text{opt}}$ is infeasible because the function $g_n$ is unknown in practice. Suppose we have an estimate $\widehat g_n$ of this function. Then, the derivation~\eqref{eq:optimal-semiparametric} motivates us to base inference on the sample covariance between $\prx$ and $\pry$ after adjusting for $\prz$:
\begin{equation}
	\widehat \rho_n(\srx, \sry, \srz) \equiv \frac1n\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)).
	\label{rho-hat}
\end{equation}
For $d = 1$, the CRT rejecting for large values of $|\widehat \rho_n|$ is an instance of the \textit{distilled CRT}  (dCRT) \cite{Liu2020}, proposed for its computational speed: If $\widehat g_n$ is fit only on $(\sry, \srz)$, then it need not be refit upon resampling $\widetilde X$. The above reasoning suggests that the dCRT is also a natural test to study in the context of power against semiparametric alternatives~\eqref{eq:linearity}. In preparation to study this test's power, we extend it to $d > 1$ and propose an asymptotically equivalent test that is easier to understand.

\subsection{A second-order approximation to the dCRT} \label{sec:second-order-approx}

Let us consider first the special case
\begin{equation}
\mathcal L_n(\prx | \prz) = N(\mu_n(\prz), \Sigma_n(\prz)).
\label{eq:normal-conditional}
\end{equation}
In this case, the resampling distribution of $\widehat \rho_n$ can be computed in closed form \cite{Liu2020}:
\begin{equation}
\mathcal L_n(\sqrt n \cdot \widehat \rho_n(\srxk, \sry, \srz) \mid \srx, \sry, \srz) = N(0, \widehat S_n^2),
\label{eq:normal-resampling-distribution}
\end{equation}
where
\begin{equation}
	\widehat S_n^2 \equiv \frac{1}{n}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))^2\Sigma_n(\srz_i).
	\label{eq:s-n-hat}
\end{equation}
When $d = 1$, the dCRT based on the statistic $T_n(\srx, \sry, \srz) = |\sqrt n \cdot \widehat \rho_n(\srx, \sry, \srz)|$ therefore rejects when $T_n(\srx, \sry, \srz) > \widehat S_n \cdot z_{1-\alpha/2}$, requiring no resampling. To extend this to $d > 1$, consider the standardized quantity
\begin{equation}
U_n(\srx, \sry, \srz) \equiv \widehat S_n^{-1} \sqrt{n} \widehat \rho_n = \frac{\widehat S_n^{-1}}{\sqrt{n}}\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)) \in \mathbb R^d.
\label{u-hat}
\end{equation}
It is natural to use as a test statistic the squared norm of $U_n$:
\begin{equation}
	T_n(\srx, \sry, \srz) \equiv \|U_n(\srx, \sry, \srz)\|^2.
	\label{t-n}
\end{equation}
Then, the normal resampling distribution~\eqref{eq:normal-resampling-distribution} implies that
\begin{equation}
\mathcal L_n(T_n(\srxk, \sry, \srz)|\srx, \sry, \srz) = \chi^2_d.
\end{equation}
It follows that the dCRT based on test statistic $T_n(\srx, \sry, \srz)$ yields the test
\begin{equation}
\phi^{N(\mu_n, \Sigma_n)}_n(\srx, \sry, \srz) \equiv \mathbbm 1(T_n(\srx, \sry, \srz) > c_{d,1-\alpha}),
\label{eq:crt-closed-form}
\end{equation}
where we recall that $c_{d,1-\alpha}$ is defined as the $1-\alpha$ quantile of $\chi^2_d$.

If the conditional distribution $\mathcal L_n(\prx|\prz)$ is not Gaussian, then the dCRT $\phi_n^{\mathcal L_n}$ based on $T_n(\srx, \sry, \srz)$ will not reduce to the closed-form expression~\eqref{eq:crt-closed-form}. However, we can think of the test $\phi^{N(\mu_n, \Sigma_n)}_n$ as a kind of second-order approximation for $\phi_n^{\mathcal L_n}$ as long as $\mathcal L_n(\prx|\prz)$ has first and second moments given by $\mu_n(\prz)$ and $\Sigma_n(\prz)$, respectively. Indeed, it is easy to check that the resampling distribution $\mathcal L_n(\sqrt n \cdot \widehat \rho_n(\srxk, \sry, \srz) \mid \srx, \sry, \srz)$ matches that derived in the normal case~\eqref{eq:normal-resampling-distribution} up to two moments. Under a few assumptions, we can make this intuition precise by showing that $\phi_n^{\mathcal L_n}$ is asymptotically equivalent to $\phi^{N(\mu_n, \Sigma_n)}_n$ (Theorem~\ref{thm:equivalence} below). We require the distribution $\mathcal L_n$ to satisfy the following moment conditions for fixed $c_1, c_2 > 0$:
\small
\begin{equation}
\mathcal L_n \in \mathscr L_n(c_1, c_2) \equiv \{\mathcal L_n: \|S_{n}^{-1}\| \leq c_1, \mathbb E_{\mathcal L_n}\left[(\pry - \widehat g_n(\prz))^{4} \mathbb E_{\mathcal L_n}[\|\prx - \mu_n(\prz)\|^{4}|\prz]\right] \leq c_2 \},
\label{eq:moment-conditions}
\end{equation} 
\normalsize
where
\begin{equation}
S_n^2 \equiv \mathbb E[\widehat S_n^2] = \mathbb E_{\mathcal L_n}\left[(\pry - \widehat g_n(\prz))^2 \Sigma_n(\prz)\right].
\end{equation}
Furthermore, to avoid technical complications, we assume that the estimate $\widehat g_n$ is trained independent dataset (whose size can vary arbitrarily with $n$ and is not included in the sample size $n$ used for testing). These training sets across $n$ and resulting estimates $\widehat g_n$ remain fixed throughout.

\begin{theorem} \label{thm:equivalence}
Suppose that for each $n$, $\mathcal L_n$ is a law whose first and second conditional moments are given by $\mu_n(\prz)$ and $\Sigma_n(\prz)$~\eqref{eq:conditional-mean-variance}, which satisfies the moment conditions~\eqref{eq:moment-conditions} for fixed for some $c_1, c_2 > 0$. Let $\phi_n^{\mathcal L_n}$ be the dCRT based on the test statistic $T_n(\srx, \sry, \srz)$~\eqref{eq:s-n-hat}, \eqref{u-hat}, \eqref{t-n}, with $\widehat g_n$ trained out of sample. The threshold $C_n(\sry, \srz)$ of this test~\eqref{upper-quantile} converges in probability to the $\chi^2_d$ quantile:
	\begin{equation}
		C_n(Y,Z) \overset{\mathcal L_n}\rightarrow_p c_{d,1-\alpha}.
		\label{eq:threshold-convergence}
	\end{equation}
	Furthermore, if $T_n(\srx, \sry, \srz)$ does not accumulate near $c_{d,1-\alpha}$, i.e.
	\begin{equation}
		\lim_{\delta \rightarrow 0}\limsup_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[|T_n(\srx, \sry, \srz)-c_{d,1-\alpha}| \leq \delta] = 0,
		\label{eq:non-accumulation}
	\end{equation}
	then the dCRT $\phi_n^{\mathcal L_n}$ is asymptotically equivalent to its second order approximation $\phi^{N(\mu_n, \Sigma_n)}_n$:
	\begin{equation}
		\lim_{n \rightarrow \infty}\ \mathbb P_{\mathcal L_n}[\phi^{\mathcal L_n}_n(\srx, \sry, \srz) \neq \phi^{N(\mu_n, \Sigma_n)}_n(\srx, \sry, \srz)] = 0.
		\label{eq:asymptotic-equivalence}
	\end{equation}	
\end{theorem}
